---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages and data

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(knitr)
library(scales)
library(formattable)
### Load data
load("movies.Rdata")
```

* * *

## Part 1: Data

The goal of this section is: 
*Describe how the observations in the sample are collected and the implications of this data collection method on the scope of inference (generalizability / causality).*

###Introduction to the Data

The dataset "movies" is comprised of 651 **randomly sampled movies** produced and released before 2016. The movies are from American Studios.  (We know this because the MPAA Ratings in teh data applies only to American movies.)  This dataset includes information from both Rotten Tomatoes and IMDb.

####About Rotten Tomatoes 
(see https://www.rottentomatoes.com/about/):

Rotten Tomatoes and the Tomatometer™ rating is the most trusted measurement of quality entertainment. As the leading online aggregator of movie and TV show **reviews from professional critics**, Rotten Tomatoes offers the most comprehensive guide to what's fresh. The world famous Tomatometer™ rating represents the percentage of positive professional reviews for films and TV shows and is used by millions every day, to help with their entertainment viewing decisions. Rotten Tomatoes designates the best reviewed movies and TV shows as Certified Fresh. That accolade is awarded with Tomatometer ratings of 75% and higher and a required minimum number of reviews. 

####About IMDB 
(see https://en.wikipedia.org/wiki/IMDb)

The Internet Movie Database (abbreviated IMDb) is an online database of information related to films, television programs and video games, including cast, production crew, fictional characters, biographies, plot summaries, trivia and reviews. As of June 2016, IMDb has approximately 3.7 million titles (including episodes) and 7 million personalities in its database.

The site enables registered users to submit new material and request edits to existing entries. Although all data is checked before going live, the **system has been open to abuse and occasional errors are acknowledged**. Users are also invited to rate any film on a scale of 1 to 10, and the totals are converted into a weighted mean-rating that is displayed beside each title, with online filters employed to deter ballot-stuffing. The site also features message boards which stimulate regular debates among authenticated users.

###Scope of Inference
In *Introduction to the Data* above, a few pharases have been **highlighted**.  These phrases provide key information about the generalizability of the data.  This information suggests the dataset should be considered the result of an observational retrospective study that uses a random sampling design to select a representative sample from U.S. movies. When random sampling has been employed in data collection, the results should be generalizable to the target population. Therefore, the results of the analysis should be generailizable to all the movies relased between 1970 - 2014.

Note that observational studies show associations. In a data analysis, association does not imply causation. Causation can only be inferred from a randomized experiment.  This analysis does not meet the requirements of an experiment.  

Sources of Bias

- “Face-to-face interviews” performed at the University of Chicago introduce a “convenience bias” as people close to Chicago may be more likely to participate.
- The study may suffer from “voluntary response bias” since people with strong responses participate. The voluntary participants may not be representative of the US population.

Non-Response Bias:  Simple random sampling means that each case in the population has an equal chance of being included and there is no implied connection between the cases in the sample. However, even when people are picked at random, caution
must be exercised if the non-response is high. If only 30% of the people randomly sampled for a survey respond, then it is uncertain whether the results are representative of the entire population. This non-response bias can skew results.

Voluntary Response Bias:  Voluntary response bias occurs when sample members are self-selected volunteers, as in voluntary samples.  Since volunteers are involved with the collection of movie data, this may be a potential souece of bias that could affect results.

* * *

## Part 2: Research question

The project objective for this section:

*Develop a research question that you want to answer using these data and a multiple linear regression model. You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.  Include a brief discussion why this question is of interest to you and/or your audience.*

I have a teenage daughter who sees movies with her friends every week.  Occasionally I take a gaggle of girls to the movies and see it with them - sitting in a different part of the theater of course!  Some of the movies are just terrtible - in my opinion.  I am often amazed how frequently my perceptions differ from young teenagers.  (I know I should not be surprised.)

Research Question:  What does it take to make a movie popular?  

I am really curious.  Do the variables that contribute to movie popularity more closely match my views or the perceptions of a teenage?

* * *

## Part 3: Exploratory data analysis

The project objective for this section:

*Perform exploratory data analysis (EDA) that addresses the research question. The EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation.*

### Data details

Before any data analysis, understanding the data is critical.  Lets first look at the 32 variables available to us:

The data frame has 651 records and 32 attributes:

- `title`: Title of movie
- `title_type`: Type of movie (Documentary, Feature Film, TV Movie)
- `genre`: Genre of movie (Action & Adventure, Comedy, Documentary, Drama, Horror, Mystery & Suspense, Other)
- `runtime`: Runtime of movie (in minutes)
- `mpaa_rating`: MPAA rating of the movie (G, PG, PG-13, R, Unrated)
- `studio`: Studio that produced the movie
- `thtr_rel_year`: Year the movie is released
- `thtr_rel_month`: Month the movie is released
- `thtr_rel_day`: Day of month the movie is released
- `dvd_rel_year`: Year the DVD is released
- `dvd_rel_month`: Month the DVD is released
- `dvd_rel_day`: Day od month the DVD is released
- `imdb_rating`: The IMDB rating
- `imdb_num_votes`: Number of votes on IMDB
- `critics_rating`: Categorical variable for critics rating on Rotten Tomatoes (Certified Fresh, Fresh, Rotten)
- `critics_score`: Critics score on Rotten Tomatoes
- `audience_rating`: Categorical variable for audience rating on Rotten Tomatoes (Spilled, Upright)
- `audience_score`: Audience score on Rotten Tomatoes
- `best_pic_nom`: Whether or not the movie was nominated for a best picture Oscar (no, yes)
- `best_pic_win`: Whether or not the movie won a best picture Oscar (no, yes)
- `best_actor_win`: Whether or not one of the main actors in the movie ever won an Oscar (no, yes) -- note that this is not necessarily whether the actor won an Oscar for their role in the given movie
- `best_actress win`: Whether or not one of the main actresses in the movie ever won an Oscar (no, yes) -- not that this is not necessarily whether the actresses won an Oscar for their role in the given movie
- `best_dir_win`: Whether or not the director of the movie ever won an Oscar (no, yes) -- not that this is not necessarily whether the director won an Oscar for the given movie
- `top200_box`: Whether or not the movie is in the Top 200 Box Office list on BoxOfficeMojo (no, yes)
- `director`: Director of the movie
- `actor1-actor5`: List of first 5 main actors in the movie (abridged cast), this information was used to determine whether the movie casts an actor or actress who won a best actor or actress Oscar
- `imdb_url`: Link to IMDB page for the movie
- `imdb_id`: IMDB ID of the movie
- `rt_url`: Link to Rotten Tomatoes page for the movie

```{r}
str(movies)
```
The `audience_score` is the response variable for my analysis.  It is the variable we will solve for in the analysis.  Why did I chose `audience_score` instead of the `imdb_rating`?  No specific reason.  It was a coin toss.  Both variables are highly correlated so picking one over the other is a random choice on my part.  Here is the correlation calculation:
```{r correlation}
cor(movies$imdb_rating, movies$audience_score)
```

Not all of the remaining variables are interesting to answer the research question.  I will also intentionally remove a few variables.  These are variables that I have determined are not relevant to my research question.   I recognizie this is a subjective decision on my part.  I also remove the IMDB rating varibles because of their high corelation to the `audience_score`:
```{r selectVariables}
movies_filtered <- select(movies, c(-(thtr_rel_year:dvd_rel_day), -top200_box, -imdb_rating, -imdb_num_votes, -(actor1:rt_url)))
```
Because my question surronds movie popularity with an interest in understanding if the result more closely mirrow my perception or my daughter's perception, I will also filter the records removing Documentaries - my daughter is not going to watch one of those!  Also, I am specifically interested in movies in theaters - not TV movies.  Because my daughter is a young teenage, I also removed all morivies rated more mature than PG-13.  Lastly, I want only complete records.
```{r}
movies_filtered <- filter(movies_filtered, title_type == "Feature Film")
movieRating <- c("G", "PG", "PG-13")
movies_filtered <- filter(movies_filtered, mpaa_rating %in% movieRating)
movies_filtered <- na.omit(movies_filtered)
```
There are still some variables that I do not need for my analysis.  For example, I have make the assertion that the name of the movie is not relevant.  Similarly, I do no tneed the `title_type` since I have already selected *Feature Film*.  The varible `critics_rating` is simply a categorical representation of `critics_score`, it too is removed from the data set.  Simliarly, `audience_rating` is a categorical representation of `audience_score` si it too is removed.  Lastly, I have asked my daughter and her friends if the studio that makes a movie makes any difference to them.  The answer was a resounding *No*.  I received the same response when I asked about the director of a movie.
```{r}
movies_filtered <- select(movies_filtered, c(-(title:title_type), -critics_rating, -audience_rating,-studio, -director))
```
Let's see what our reduced data looks like:
```{r}
dim(movies_filtered)
str(movies_filtered)
```


```{r}
ggplot(movies_filtered, aes(x = audience_score)) + geom_histogram() + xlab("Audience Score") + ylab("Count") + ggtitle("Histogram of Audience Score")
summary(movies_filtered$audience_score)
```
The histogram does appear a bit left skewed but this is understnabale.  I submit that popular movies typcally get rated highly by a broad audience where a less popular movies will always have a measureable following.

```{r}
ggplot(movies_filtered, aes(x = title_type, y = audience_score)) + geom_boxplot() + xlab("Title Type") + ylab("Audience Score") + ggtitle("Audience Score vs. Title Type")
movies_filtered %>% group_by(title_type) %>% summarise(median_score = median(audience_score), iqr_score = IQR(audience_score))
```

```{r uniqueValues}
unique1 <- lapply(movies_filtered, unique)
unique2 <- sapply(unique1, length)
unique2 <- as.data.frame(unique2)
colnames(unique2) <- "Count_Distinct"
unique3<-cbind(Variables=rownames(unique2), unique2)
rownames(unique3) <- NULL
unique4 <- unique3 %>% mutate(Percent_Unique = 1-Count_Distinct/nrow(movies_filtered))
unique4$Percent_Unique <- percent(unique4$Percent_Unique)
format_table(unique4,list(Variables=formatter("span", style="color:blue")), align = "c")
kable(unique4)
ggplot(unique4, aes(x=Variables, y=Percent_Unique)) + geom_bar(stat = "identity")
```

http://www.cookbook-r.com/Manipulating_data/Summarizing_data/


```{r evalDirectors}
by_directors <- movies_filtered %>% group_by(director) %>% summarize(n()) 
colnames(by_directors) <- c("Director_Name", "Count_Movies")
by_directors <- by_directors[order(-by_directors$Count_Movies),]
range(by_directors$Count_Movies)
by_dir_cnt <- by_directors %>% group_by(Count_Movies) %>% summarize(n())
colnames(by_dir_cnt) <- c("Num_Movies", "Num_Directors")
ggplot(by_dir_cnt, aes(x=Num_Movies, y=Num_Directors)) + 
     geom_bar(stat = "identity", fill="lightgreen") + 
     geom_text(aes(label=Num_Directors), vjust=-.5) +
     ggtitle("How Many Movies Did Directors Produce?") +
     labs(x="No. of Movies Produced", y="No. of Directors")
```

     
* * *

## Part 4: Modeling

The project objective for this section:

*Develop a multiple linear regression model to predict a numerical variable in the dataset. The response variable and the explanatory variables can be existing variables in the dataset, or new variables you create based on existing variables.*

* * *

## Part 5: Prediction

The project objective for this section:

*Pick a movie from 2016 (a new movie that is not in the sample) and do a prediction for this movie using your the model you developed and the `predict` function in R. Also quantify the uncertainty around this prediction using an appropriate interval.*


* * *

## Part 6: Conclusion

The project objective for this section:

*A brief summary of your findings from the previous sections without repeating your statements from earlier as well as a discussion of what you have learned about the data and your research question. You should also discuss any shortcomings of your current study (either due to data collection or methodology) and include ideas for possible future research.*
